%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -*- Mode: Latex -*- %%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% intro.tex --- 
%%% Author          : blehr
%%% Created On      : Mon Jan 18 10:34:12 1993
%%% Last Modified By: blehr
%%% Last Modified On: Sun Mar 21 07:45:51 1993
%%% RCS revision    : $Revision$ $Locker$
%%% Status          : In writing
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}
\label{intro}

\section{Synchronization and Cognition}
\label{intro:synch}

{\em Synchronization and Cognition\/} (SC) is a cooperation project
initiated by R.\ Eckhorn and R.\ Bauer of the Biophysics Group and F.\ 
R\"{o}sler of the Physiological Psychology Group at the
Philipps-Universit\"{a}t Marburg, Germany.

The basis of the current work is observations done by R.\ Eckhorn and
R.\ Bauer~(\cite{cat}), as well as by the group of W.\ Singer
(e.g.,~\cite{singer2} and~\cite{singer1}), that when stimulated,
retinally distributed and rhythmically correlated neural assemblies
organize in peripheral visual areas in the brain of drugged cats.  The
spatial distribution of these assemblies, but not their frequencies,
depend on the supplied visual stimulus.

The goal of the SC project is, through the use of microelectrodes
(monkeys) and EEG recordings (humans), to show that coherent
neurological activations ({\em synchronizations\/}) in the visual
cortex, relating to coherent visual stimuli, is a requirement for the
coherent perception of these stimuli.  In addition, the time-space
{\em synchronization dynamics\/} in peripheral and central, as well as
in intermediate cortex areas, are to be investigated as to how they
relate to the cognitive tasks being performed.

For a further description of the SC project, see~\cite{antrag}.  Some
of the publications by R.\ Eckhorn {\em et al.\/} concerning the
discovery and implications of synchronizations in the visual cortex of
cats and monkeys are~\cite{eckhorn1},~\cite{eckhorn2},~\cite{cat}
and~\cite{eckhorn3}.

The only group of the SC project with which I have interacted during
my thesis work, is the group of R.\ Eckhorn and R.\ Bauer.
Accordingly, I will in the following refer to the {\em monkey\/} and
the {\em subject\/} interchangeably.  The general principles, however,
when referring to the monkey, can be assumed also to apply to the work
done by the group of F.\ R\"{o}sler.

\section{Thesis Motivation}
\label{intro:motivation}

Since the direction of gaze and the orientation of the visual stimuli
have to correspond as exactly as possible, in order to be certain that
exactly those neural assemblies are stimulated whose activity is being
registered, it is of vital importance for the validity of the
experimental results that the exact eye position of the subject be
known at a given time.  Otherwise the danger exists of registering
neurological responses whose variations are not the result of the
assigned cognitive task, but of the subject's incorrect gaze
direction.  

In the part of the SC project carried out by the Eckhorn group, the
visual stimuli are generated on a computer monitor placed in front of
the monkey, sometimes referred to as the {\em inducing monitor\/}
(this term stems from the fact that the neural activity in the visual
cortex of the monkey is {\em induced\/} by the visual patterns on the
monitor).  The monkey has been trained to focus on a bright point,
called the reference point, on the monitor, across which various
geometrical shapes move in different directions.  When the monkey does
not focus on the reference point, it is said to be {\em drifting\/}.

To gain absolute certainty about where the monkey at a given time
focuses, it is necessary to have a system which continuously tracks
the movements of the monkey's eyes.  For our purpose it was decided
that a system complying to the following requirements would suffice:
\begin{enumerate}
\item The position-analyzing method of the system may in no way
  interfere with recordings being made from the visual cortex of the
  monkey.
\item\label{req:first}A position measuring rate of 50 Hz.  Thus every
  20 ms, hereafter referred to as a {\em pass\/}, a coordinate pair
  relative to a reference point on the inducing monitor, corresponding
  to the point on the monitor on which the monkey had its focus at the
  beginning of the pass, has to be made available.  When the monkey is
  drifting, a warning signal has to be given with each pass as long as
  it continues to drift.
\item\label{req:2} The error in the position returned by the system
  should be less than or equal to $\pm 0.25^{\circ}$.
\item\label{req:3}The measuring range of the system should be at least
  $\pm 20^{\circ}$.
\item\label{req:4}The system should be relatively easy to configure to
  comply with the specific needs of an experiment.
\item\label{req:last}The price of the complete system may not exceed
  DM 12.000.
\end{enumerate}

There exist several commercially available eye tracking systems, some
of which are described in Section~\ref{back:track}.  For reasons
elaborated upon in Section~\ref{back:track:summary}, it was decided to
develop an entirely new eye tracking system, based on real-time image
processing of video input.  And this is, in short, what has been the
subject of my thesis; to develop, and, as far as possible, test a
suitable {\em software platform\/} for a real-time eye position
measuring system complying to the requirements of precision and speed
mentioned above.
